\section*{Acknowledgments}

I thank my co-advisor \firstAdvisorName, for his extraordinary guidance and assistance throughout the Master Degree, for taking me into the wonderful world of Bioinformatics, for his dedication to this project, and for giving me a thesis project that I am truly passionate about. This thesis would not exist without him.\\

I thank my co-advisor \secondAdvisorName, for teaching me so much about Genetics, Statistics and Quality Control, for giving me very valuable feedback and pointers to guide me, and for his constant dedication to this project. This thesis would not be as thorough nor clean without his assistance.\\

I thank Dr. José Gerardo Taméz Peña, for the development of the FRESA.CAD Benchmark and his assistance in applying the benchmark to the datasets, and for the work we have developed together.\\

I thank Dr. Claudia Marquez-Luna, for the development of the LDPred-funct method and for her assistance in applying it to the datasets.\\

I thank the committee members \thirdAdvisorName  and \fourthAdvisorName, for taking the time to provide valuable time in the thesis defense procedure and for their comments and feedback to make this thesis as complete and correct as possible.\\

I thank my degree Co-ordinator \programDirectorName, for providing support and guidance across the Masters Degree and further career choices.\\

I thank the Tecnologico de Monterrey for the resources and infrastructure provided that supported the development of this thesis. I also thank it for the scholarship provided to cover tuition fees across the degree period.\\

I thank the Consejo Nacional de Ciencia y Tecnología (CONACYT) for the scholarship and funding during the period I dedicated to the Masters Degree, and in general for supporting the development of science in Mexico.\\

I thank our colleagues from the Bioinformatics for Clinical Diagnosis Research Program, School of Medicine and Health Sciences, Tecnologico de Monterrey, for their valuable comments on this work. \\


I thank the Alzheimer's Disease Neuroimaging Initiative for providing the Whole-Genome Sequence data used in this project. Data collection and sharing for this project was funded by the Alzheimer's Disease Neuroimaging Initiative(ADNI) (National Institutes of Health Grant U01 AG024904) and DOD ADNI (Department of Defense award number W81XWH-12-2-0012). ADNI is funded by the National Institute on Aging, the National Institute of Biomedical Imaging and Bioengineering, and through generous contributions from the following: AbbVie, Alzheimer’s Association; Alzheimer’s Drug Discovery Foundation; Araclon Biotech; BioClinica, Inc.; Biogen; Bristol-Myers Squibb Company; CereSpir, Inc.; Cogstate; Eisai Inc.; Elan Pharmaceuticals, Inc.; Eli Lilly and
Company; EuroImmun; F. Hoffmann-La Roche Ltd and its affiliated company Genentech, Inc.; Fujirebio; GE Healthcare; IXICO Ltd.; Janssen Alzheimer Immunotherapy Research \& Development, LLC.; Johnson \& Johnson Pharmaceutical Research \& Development LLC.; Lumosity; Lundbeck; Merck \& Co., Inc.; Meso Scale Diagnostics, LLC.; NeuroRx Research; Neurotrack Technologies; Novartis Pharmaceuticals Corporation; Pfizer Inc.; Piramal Imaging; Servier; Takeda Pharmaceutical Company; and Transition Therapeutics. The Canadian Institutes of Health Research is providing funds to support ADNI clinical sites in Canada. Private sector contributions are facilitated by the Foundation for the National Institutes of Health (www.fnih.org). The grantee organization is the Northern California Institute for Research and Education, and the study is coordinated by the Alzheimer’s Therapeutic Research Institute at the University of Southern California. ADNI data are disseminated by the Laboratory for Neuro Imaging at the University of Southern California.\\

I thank the International Genomics of Alzheimer's Project (IGAP) for providing summary results data for these analyses. The investigators within IGAP contributed to the design and implementation of IGAP and/or provided data but did not participate in analysis or writing of this report. IGAP was made possible by the generous participation of the control subjects, the patients, and their families. The i–Select chips was funded by the French National Foundation on Alzheimer's disease and related disorders. EADI was supported by the LABEX (laboratory of excellence program investment for the future) DISTALZ grant, Inserm, Institut Pasteur de Lille, Université de Lille 2 and the Lille University Hospital. GERAD was supported by the Medical Research Council (Grant n 503480), Alzheimer's Research UK (Grant n 503176), the Wellcome Trust (Grant n 082604/2/07/Z) and German Federal Ministry of Education and Research (BMBF): Competence Network Dementia (CND) grant n 01GI0102, 01GI0711, 01GI0420. CHARGE was partly supported by the NIH/NIA grant R01 AG033193 and the NIA AG081220 and AGES contract N01–AG–12100, the NHLBI grant R01 HL105756, the Icelandic Heart Association, and the Erasmus Medical Center and Erasmus University. ADGC was supported by the NIH/NIA grants: U01 AG032984, U24 AG021886, U01 AG016976, and the Alzheimer's Association grant ADGC–10–196728.





%%%%%%%%%%%%%
\begin{center}
\end{center}
%%%%%%%%%%%%%

\begin{flushright}  
	\MakeUppercase{\authorName}
\end{flushright}  

\begin{center}
    \textbf{\schoolName}\\
    \textbf{\schoolPlace}, \textbf{\thesisDate} 
\end{center}
